---
title: "Sediment accumulation rates - DataPrep"
output: html_notebook
---

# Preparations

## Install packages

```{r packages, message=FALSE, warning=FALSE}
rm(list=ls())

library(sf)
library(sdmpredictors)
library(spatialEco)
library(dplyr)
library(stars)
library(terra)
```


## Define projection and resolution

Projection based on https://projectionwizard.org/ using the AoI.

```{r projection}
crs <- "+proj=laea +lat_0=90 +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs" 
res <- 4000
```


## Define Area of Interest (AoI)

The area of interest is defined by the predicted sediment classes. Rocks and boulders (50) define areas outside the AoI.

```{r aoi}
AoI <- rast("input/GrainSizeReg_folk8_classes_2023-06-28.tif")
AoI[AoI == 11] <- 1
AoI[AoI == 12] <- 1
AoI[AoI == 13] <- 1
AoI[AoI == 20] <- 1
AoI[AoI == 30] <- 1
AoI[AoI == 40] <- 1
AoI[AoI == 50] <- NA
AoI[AoI == 60] <- 1

AoI <- as.polygons(AoI, dissolve = TRUE)
```


# Predictor variables

Potentially relevant predictor variables are loaded and a raster stack is created.

```{r load_predictors}
predictors <- rast("input/predictors_ngb.tif")
mud <- extend(rast("input/mud_2023-06-30.tif"), predictors)

sedenv_prob <- extend(rast("input/SedEnv3_probabilities2023-07-01.tif"), predictors)

sedenv_maxprob <- extend(rast("input/SedEnv3_max_probabilities2023-07-01.tif"), predictors)

sedenv_relprob <- sedenv_prob / sedenv_maxprob

predictors <- c(predictors, mud, sedenv_relprob)
names(predictors)[38] <- "MUD"
names(predictors)[39] <- "DEPOrelprob"
names(predictors)[40] <- "EROrelprob"
names(predictors)[41] <- "NDEPrelprob"
```


## Ensure uniform projection

Check if AoI and predictors have the defined projection. Re-project if this is not the case.

```{r uniform_proj}
if (st_crs(AoI)$proj4string != crs) {
  AoI <- st_transform(AoI, crs)
}

if (crs(predictors) != crs) {
  predictors <- project(predictors, crs, res = res)
}
```


## Crop predictors to AoI

```{r crop_predictors}
predictors <- crop(mask(predictors, AoI, touches = FALSE), AoI)
plot(predictors)
```


## Create a fishnet based on predictor raster

```{r fishnet}
min_ext <- sum(predictors)
min_ext[min_ext == 0] <- NA
fishnet <- st_as_sf(stars::st_as_stars(min_ext), as_points = FALSE, merge = FALSE)
fishnet$ID <- 1:nrow(fishnet)
```


# Response variable

## Load response

The dataset is restricted to sedimentation rates measured with 210Pb. Two data points are removed as measured sedimentation rates are possibly too high as a result of bioturbation (de Haas et al., 1997).

```{r load_response}
resp_data <- read.csv("input/norway_sar_2023-08-28.csv", header = TRUE, sep = ",")
resp_data <- subset(resp_data,  SAR_cm_yr_method == 'Pb-210'| SAR_cm_yr_method == 'Pb-210 CFCS' | SAR_cm_yr_method == 'Pb-210 CIC' | SAR_cm_yr_method == 'Pb-210 CRS' | SAR_cm_yr_method == 'Pb-210 Cs-137' | SAR_cm_yr_method == 'Pb-210 Cs-137 CFCS') # Retain only sedimentation rates based on Pb-210
resp_data <- subset(resp_data, core_id != 12379 & core_id != 12380) # Rates possibly too high due to bioturbation
resp <- resp_data[,c(3,1,2,22)] #Reduce to required columns
summary(resp)
```


## Convert to spatial object

```{r convert_to_spatial}
resp <- st_as_sf(resp, coords = c("longitude", "latitude"), crs = "+proj=longlat +datum=WGS84 +no_defs")
```


## Ensure uniform projection

Check if the response variable has the defined projection. Re-project if this is not the case.

```{r uniform_proj_oc}
if (st_crs(resp)$proj4string != crs) {
  resp <- st_transform(resp, crs)
}
```


## Clip to minimum extent

```{r clip_response}
resp <- st_intersection(resp, st_as_sf(as.polygons(min_ext, dissolve = TRUE)))
```


## Pseudo-samples

To account for areas where erosion dominates, pseudo-samples with a sedimentation rate of 0 cm/yr will be created. The creation of pseudo-samples is based on the spatial model of the sedimentary environment. To increase the likelihood of a correct assignment, pseudo-samples will be restricted to erosional areas with a relative probability of equal to or larger than 0.99 and within the area of applicability.

### Create a mask

First, a mask is created. Samples will be randomly assigned within the mask.

```{r create_mask}
# Matrix to reclassify relative erosion probabilities
m <- c(0, 0.99, 0, 0.99, 1, 1)
rcl_matrix <- matrix(m, ncol = 3, byrow = TRUE)

# Reclassify
e99 <- classify(predictors$EROrelprob, rcl_matrix, right = FALSE)

# Load AOA
aoa <- rast("N:/Prosjekter/311700_MAREANO/311778_Automatisert_kartlegging/Sedimentasjon/R/SedEnv/output/SedEnv3_AOA2023-07-01.tif")

mask <- e99 * aoa
plot(mask)
```


### Calculate number of pseudo-samples

The number of pseudo-samples to be created will be proportional to the size of the mask, in which the samples are created.

```{r number_pseudo-samples}
fr <- freq(mask)
f <- fr$count[2]/sum(fr$count)

n <- round(f/(1-f)*nrow(resp))
```


### Create pseudo-samples

Pseudo-samples are then created randomly.

```{r create_pseudo_samples}
mask[mask == 0] <- NA

set.seed(42)
pseudo_smp <- spatSample(mask, size = n, method = "random", na.rm = TRUE, as.points = TRUE)

pseudo_smp <- st_as_sf(pseudo_smp)
names(pseudo_smp)[1] <- "sum"
pseudo_smp$core_id <- "pseudo_sample"
pseudo_smp$SAR_cm_yr <- 0
pseudo_smp <- pseudo_smp[,c(3,4,1,2)]
```


## Combine both datasets

```{r combine}
resp <- rbind(resp, pseudo_smp)
```


## Centroids

```{r centroids, warning=FALSE}
centroids <- st_join(fishnet, resp)
centroids <- na.omit(centroids)
centroids <- centroids %>% group_by(ID) %>% summarize(SAR_cm_yr = mean(SAR_cm_yr, na.rm = TRUE))
centroids <- st_centroid(centroids)

resp_centr <- centroids
```


# Save files for use in main process

```{r save_predictors}
write_sf(resp_centr, "input/SAR.shp", overwrite = TRUE)
writeRaster(predictors, "input/predictors.tif", overwrite = TRUE)
```


## Save session info

```{r save_session_info}
sessionInfo <- sessionInfo()
save(sessionInfo, file = "sessionInfo_DataPrep.Rdata")
rm("sessionInfo")
```


## Save global environment

```{r save_global_env}
save.image(file = "globEnv_DataPrep.RData")
```

