---
title: "Sediment accumulation Rates - Main"
output:
  html_document:
    df_print: paged
---

# Introduction

210Pb Sedimentation rates of surface sediments on the Norwegian continental margin are spatially predicted based on observations (response variable) and predictor variables using quantile regression forests. 

# Preparations

## Install packages

```{r packages, message=FALSE, warning=FALSE}
rm(list=ls())

library(terra)
library(dplyr)
library(caret)
library(ggplot2)
library(sf)
library(CAST)
library(lwgeom)
library(geosphere)
library(quantregForest)
library(doParallel)
library(ModelMetrics)
library(forcats)
```


## Load required data

```{r load_data}
predictors <- rast("input/predictors.tif")
resp_centr <- vect("input/SAR.shp")
AoI <- read_sf("input/AoI.shp")

names(predictors)
```


## Type of response?

Define which property is used as response data.

```{r response_type}
resp_type <- "SAR"
resp_unit <- "cm/yr"
```


## Plot data

```{r plot_data}
plot(predictors$BATH)
plot(resp_centr, pch = 20, col = "black", cex = 0.5, add = TRUE)
```


## Create a regression matrix

```{r regression_matrix}
ov_resp <- as.data.frame(extract(predictors, resp_centr))

rm_resp <- cbind(resp_centr$SAR_cm_yr, ov_resp)
rm_resp <- rm_resp[,-2]
names(rm_resp)[1] <- resp_type

summary(rm_resp)
```


# Data exploration

## SAR versus water depth

```{r sar_wdepth}
plot(rm_resp$SAR, rm_resp$BATH, xlab = paste0(resp_type, " (", resp_unit, ")"), ylab = "Water depth (m)")
```


## SAR histogram

```{r hist_sar}
hist(rm_resp$SAR, main = "", xlab = paste0(resp_type, " (", resp_unit, ")"))
```


## Distances in geographic space

Euclidean nearest neighbour distances in geographic space are calculated and visualized. This type of visualization allows to assess whether training data feature a representative coverage of the prediction area and if cross-validation (CV) folds are adequately chosen to be representative for the prediction locations. See e.g. Meyer and Pebesma (2022) and Milà et al. (2022) for further discussion on this topic.

```{r geogr_space_dist, message=FALSE}
dist_geogr <- plot_geodist(st_as_sf(resp_centr), predictors,
                     type = "geo",
                     unit = "km",
                     showPlot = FALSE)

dist_geogr$plot
dist_geogr$plot + scale_x_log10()
```


# Quantile Regression Forest model

## Creating spatial blocks

Spatial blocks and folds are created. The folds will be used in a spatial k-fold cross validation. The k-fold nearest neighbour distance matching algorithm is used here.

```{r nndm}
k <- 10 # Number of folds
knndmfolds <- knndm(tpoints = st_as_sf(project(resp_centr, AoI)),
                    modeldomain = AoI,
                    k = k,
                    samplesize = 2000)
```


## Distances in geographic space including CV distances

After creating spatial blocks and folds the cross validation (CV) distances should approximate the sample-to-prediction distances in geographic space.

```{r geogr_space_dist2, message=FALSE}
dist_geogr2 <- plot_geodist(st_as_sf(resp_centr), predictors,
                     cvfolds= knndmfolds$indx_test,
                     type = "geo",
                     unit="km",
                     showPlot = FALSE)

dist_geogr2$plot
dist_geogr2$plot + scale_x_log10()
```


## Model tuning

A Quantile Regression Forest model is tuned. Predictor variables are selected in a forward feature selection approach and various values of the mtry parameter are tested in a spatial k-fold cross validation.

The maximum number of iterations to be performed can be calculated upfront, based on the number of pre-selected predictors:

```{r max_iter}
factorial(length(names(predictors)))/(factorial(2)*factorial(length(names(predictors))-2)) + sum(c((length(names(predictors))-2):1))
```


### Forward feature selection

The best combination of predictor variables (features) is found in a forward feature selection process.

```{r model_tuning}
nCores <- detectCores()
cl <- makePSOCKcluster(nCores - 1)
registerDoParallel(cl)

set.seed(42)

model <- ffs(rm_resp[names(predictors)],
               rm_resp$SAR,
               metric = "Rsquared",
               method = "qrf",
               what = 0.5,
               replace = FALSE,
               importance = TRUE,
               trControl = trainControl(method="CV",
                                        number = k,
                                        savePredictions = "final",
                                        index = knndmfolds$indx_train,
                                        allowParallel = TRUE),
               verbose = TRUE)

stopCluster(cl)

model

sel_preds <- model$selectedvars
```


### FFS plot

Plot of R2 over the model runs.

```{r ffs_plot}
plot_ffs(model)
```


## Validation statistics

The validation results of the optimal RF model.

Note that these are the statistics based on the predicted values of the selected model. These differ slightly from the values from the tuning (above), which are the means of the k predictions based on the folds.

```{r validation_stats}
t <- data.frame(model$pred$pred, model$pred$obs)

validation <- data.frame(mse=numeric(), rmse=numeric(), r2=numeric())
validation[1,1] <- round(sum(t$model.pred.obs - t$model.pred.pred)/nrow(t), 3)
validation[1,2] <- round(rmse(t$model.pred.obs, t$model.pred.pred), 3)
validation[1,3] <- round(cor(t$model.pred.obs, t$model.pred.pred)^2, 3)

colnames(validation) <- c("ME", "RMSE", "r2")
rownames(validation) <- NULL
validation
```


## Validation plot

```{r validation_plot, message=FALSE}
ggplot(t, aes(x = model.pred.pred, y = model.pred.obs)) +
  geom_point() +
  geom_smooth(method = "lm") +
  geom_abline(intercept = 0, slope = 1, colour = "grey", linewidth = 1.2) +
  theme_bw() +
  scale_x_continuous(name = "Predicted value") +
  scale_y_continuous(name = "Observed value") +
  ggtitle(paste0(resp_type, " (", resp_unit, ")"))
```


## Variable importance

```{r variable_importance_plot, warning=FALSE}
imp <- varImp(model$finalModel)
imp$Predictor <- rownames(imp)
rownames(imp) <- NULL
imp <- imp[order(imp[1], decreasing = TRUE), c(2, 1)]
colnames(imp)[2] <- "IncMSE"
imp

impfig <- imp %>%
  mutate(Predictor = fct_reorder(Predictor, IncMSE)) %>%
  ggplot( aes(x=Predictor, y=IncMSE)) +
    geom_bar(stat="identity", fill="#f68060", alpha=.6, width=.4) +
    coord_flip() +
    xlab("") +
    ylab("% increase in MSE") +
    ggtitle("Sediment accumulation rate") +
    theme_bw()
    
impfig

jpeg(filename = "output/SAR_VarImp.jpg", width = 12, height = 12, units = "cm", res = 500)
impfig
dev.off()
```


## Partial dependence

Partial dependence plots give a graphical depiction of the marginal effect of a variable on the response.

```{r partial_plots}
m2 <- model$finalModel
class(m2) <- "randomForest"

for (i in 1:length(sel_preds)) {
  partialPlot(x = m2, pred.data = rm_resp, x.var = sel_preds[i], main = "", xlab = sel_preds[i], ylab = paste0(resp_type, " (", resp_unit, ")"))
}

for (i in 1:length(sel_preds)) {
  jpeg(filename = paste0("output/", sel_preds[i], ".jpg"), width = 12, height = 12, units = "cm", res = 500)
  par(mar = c(4.5,5,1,1))
  partialPlot(x = m2, pred.data = rm_resp, x.var = sel_preds[i], ylim=c(0,0.41), main = "", xlab = sel_preds[i], ylab = expression(Sediment ~ accumulation ~ rate ~ (cm ~ yr^-1)))
  dev.off()
}
```


# Predict QRF model

## Predict SAR

```{r predict_sar}
preds <- raster::stack(predictors[[sel_preds]])
SAR_med <- predict(preds, model$finalModel, what = 0.5)
SAR_p95 <- predict(preds, model$finalModel, what = 0.95)
SAR_p5 <- predict(preds, model$finalModel, what = 0.05)
SAR_pi90 <- SAR_p95 - SAR_p5
SAR_pir <- SAR_pi90 / SAR_med

hist(SAR_med, breaks = 40, main = "", xlab = paste0(resp_type, " (", resp_unit, ")"))
```


## Area of Applicability

```{r aoa}
SAR_trainDI <- trainDI(model = model,
                        variables = sel_preds)
print(SAR_trainDI)

SAR_aoa <- aoa(newdata = predictors, 
                model = model,
                trainDI = SAR_trainDI,
                variables = sel_preds,
                )

plot(SAR_aoa)
```


## Plot results

```{r plot_results}
plot(SAR_med, main = paste0(resp_type, " (", resp_unit, ")"))
plot(SAR_pi90, main = "90% prediction interval")
plot(SAR_pir, main = "Prediction interval ratio")
plot(SAR_aoa$DI, main = "Dissimilarity index")
plot(SAR_aoa$AOA, main = "Area of applicability")

fr <- freq(SAR_aoa$AOA)
print(paste0("AOA = ", round(100*fr$count[2]/ sum(fr$count),2), "% of pixels"))
```


## Convert AOA from raster to polygon

```{r aoa_poly}
aoa_poly <- as.polygons(SAR_aoa$AOA, dissolve = TRUE)
plot(aoa_poly)

write_sf(st_as_sf(aoa_poly), dsn = "output", layer = paste0(resp_type, "_AOA_", Sys.Date()), driver = "ESRI Shapefile")
```


## Export geoTifs

```{r geotifs}
writeRaster(SAR_med, paste0("output/", resp_type, "_median_", Sys.Date(), ".tif"), overwrite = TRUE)
writeRaster(SAR_pi90, paste0("output/", resp_type, "_pi90_", Sys.Date(), ".tif"), overwrite = TRUE)
writeRaster(SAR_pir, paste0("output/", resp_type, "_pir_", Sys.Date(), ".tif"), overwrite = TRUE)
writeRaster(SAR_aoa$AOA, paste0("output/", resp_type, "_aoa_", Sys.Date(), ".tif"), overwrite = TRUE)
```


## Output a log file

```{r log}
sink(file = paste0("output/ModelLog_", Sys.Date(), ".txt"))
print("Selected Predictors")
sel_preds
model
print("Final Model")
paste0("ME = ", validation[1,1])
paste0("RMSE = ", validation[1,2])
paste0("R2 = ", validation[1,3])
paste0("AOA = ", round(100*fr$count[2]/ sum(fr$count),2), "% of pixels")
sink()
```


# Finishing off

# Save QRF model

```{r save_model}
saveRDS(model, "qrfmodel.rds")
```


## Save session info

```{r save_session_info}
sessionInfo <- sessionInfo()
save(sessionInfo, file = "sessionInfo.Rdata")
rm("sessionInfo")
```


## Save global environment

```{r save_global_env}
save.image(file = "globEnv.RData")
```
