---
title: "Sedimentation Rates - Main"
output:
  html_document:
    df_print: paged
---

# Introduction

210Pb Sedimentation rates of surface sediments on the Norwegian continental margin are spatially predicted based on observations (response variable) and predictor variables using quantile regression forests. 

# Preparations

## Install packages

```{r packages, message=FALSE, warning=FALSE}
rm(list=ls())

library(raster)
library(terra)
library(dplyr)
library(rcompanion)
library(Boruta)
library(caret)
library(usdm)
library(corrplot)
library(ggplot2)
library(sf)
library(CAST)
library(lwgeom)
library(geosphere)
library(quantregForest)
library(blockCV)
library(automap)
library(gstat)
library(doParallel)
library(ModelMetrics)
library(forcats)
```


## Load required data

```{r load_data}
predictors <- rast("input/predictors.tif")
resp_centr <- vect("input/SAR.shp")

names(predictors)
```


## Type of response?

Define which property is used as response data.

```{r response_type}
resp_type <- "SAR"
resp_unit <- "cm/yr"
```


## Plot data

```{r plot_data}
plot(predictors$BATH)
plot(resp_centr, pch = 20, col = "black", cex = 0.5, add = TRUE)
```


## Create a regression matrix

```{r regression_matrix}
ov_resp <- as.data.frame(extract(predictors, resp_centr))

rm_resp <- cbind(resp_centr$SAR_cm_yr, ov_resp)
rm_resp <- rm_resp[,-2]
names(rm_resp)[1] <- resp_type

summary(rm_resp)
```


## Data exploration

### SAR versus water depth

```{r sar_wdepth}
plot(rm_resp$SAR, rm_resp$BATH, xlab = paste0(resp_type, " (", resp_unit, ")"), ylab = "Water depth (m)")
```


### SAR

```{r hist_sar}
hist(rm_resp$SAR, main = "", xlab = paste0(resp_type, " (", resp_unit, ")"))
```


### Transformation

The response data are transformed to approach a normal distribution. This is necessary for semivariogram fitting.

Null hypothesis: The data are normally distributed. If p > 0.05, normality can be assumed.

```{r transformation}
if(nrow(rm_resp) > 5000) {
  lambda <- transformTukey(sample(rm_resp$SAR, 5000), plotit = TRUE, quiet = FALSE, returnLambda = TRUE)
  } else {
  lambda <- transformTukey(rm_resp$SAR, plotit = TRUE, quiet = FALSE, returnLambda = TRUE)
}
```


# Predictor variable selection

## Boruta algorithm

Although two models will be built, it is sufficient to run the Boruta algorithm only once, as the predictors and the sample locations are identical.

```{r boruta}
set.seed(42)
B <- Boruta(rm_resp[[1]] ~ .,data = rm_resp[2:ncol(rm_resp)], pValue = 0.05,
             maxRuns = 10000)
B
par(mar=c(13,4,1,1), cex = 0.6)
plot(B, las=2, colCode = c("greenyellow", "yellow2", "red3", "cadetblue"), xlab = "")
```


## De-correlation analysis

To reduce redundancy in information, a de-correlation analysis is carried out. Of those predictor variables identified as important in the Boruta analysis, only those with a correlation coefficient below a set threshold are retained. However, a universally applicable threshold does not exist. Additionally, multicollinearity, i.e., collinearity between three or more variables, might exist in the data. Variance inflation factors (VIFs) are therefore additionally calculated to check for multicollinearity. As a rule of thumb, VIFs larger than 5 or 10 indicate a problematic amount of collinearity (James et al., 2017: pp. 101-102; doi: 10.1080/24754269.2021.1980261). According to Johnston et al. (2017; doi: 10.1007/s11135-017-0584-6) a VIF of 2.5 or greater is generally considered indicative of considerable collinearity.

```{r de-corr, message=FALSE, warning=FALSE}
th <- 1

repeat{
 cor_result<- vifcor(rm_resp[rownames(subset(attStats(B), decision == "Confirmed"))], th = th,  maxobservations = nrow(rm_resp))
 if (max(cor_result@results[,2]) >= 2.5){
   th <- th - 0.01
 } else {
   break
 }
}

max(cor_result@results[,2])
cor_result

#sel_preds <- names(rm_resp)[2:ncol(rm_resp)]
sel_preds <- cor_result@results$Variables
seldata <- rm_resp[c(resp_type, sel_preds)]
```


##  Correlation plot

```{r correlation_plot}
corrplot.mixed(cor(rm_resp[sel_preds]), lower.col =  "black", tl.pos = "lt", number.cex = 0.6)
```


## Environmental space

A visual check to what extent the samples cover the environmental space. This is useful as legacy data were used and no formal sampling design was applied in the analysis.

* Blue: Samples

* Grey: Environmental data (based on random subsample)

```{r env_space}
smp <- as.data.frame(spatSample(x = predictors[[sel_preds]], size = nrow((rm_resp)), method = "random", na.rm = TRUE))

for (i in 2:ncol(seldata)) {
    
  print(ggplot() +
          geom_density(data = seldata, aes(x=seldata[,i]),
          colour = "cornflowerblue", fill = "cornflowerblue", alpha = 0.1,linewidth = 1) +
          geom_density(data = smp, aes(x = smp[,i-1]), colour = "grey",fill = "grey", alpha = 0.1, linewidth = 1) +
          scale_x_continuous(name = names(seldata[i])) +
          theme_bw())
        
}
```


## 2D plots of environmental space

```{r 2d_env_plots}

for (i in sel_preds[2:length(sel_preds)]) {
  
  print(ggplot() +
    geom_point(data = smp, aes(x = smp[,i], y = smp[,1]), colour = "grey", alpha = 1, size = 2) +
    geom_point(data = seldata, aes(x = seldata[,i], y = seldata[,2]), colour = "cornflowerblue", alpha = 1, size = 2) +
    scale_x_continuous(name = names(seldata[i]))+
    ylab(sel_preds[1]) +
    theme_bw())
}
```


## Distances in environmental space 

```{r env_space_dist}
#dist_env <- plot_geodist(st_as_sf(resp_centr), predictors,
#                     type = "feature",
#                     variables = sel_preds,
#                     showPlot = FALSE)

#dist_env$plot
#dist_env$plot + scale_x_log10()
```


## Distances in geographic space

```{r geogr_space_dist, message=FALSE}
dist_geogr <- plot_geodist(st_as_sf(resp_centr), predictors,
                     type = "geo",
                     unit = "km",
                     showPlot = FALSE)

dist_geogr$plot
dist_geogr$plot + scale_x_log10()
```


# Quantile Regression Forest model

## Quick RF model without spatial CV

```{r quick_rf, warning=FALSE}
set.seed(42)
rf <- randomForest(SAR ~ ., seldata, replace = FALSE)

t <- data.frame(rf$pred, rf$y)

validation <- data.frame(mse=numeric(), rmse=numeric(), r2=numeric())
validation[1,1] <- round(mse(t$rf.y, t$rf.pred), 3)
validation[1,2] <- round(rmse(t$rf.y, t$rf.pred), 3)
validation[1,3] <- round(cor(t$rf.y, t$rf.pred)^2, 3)

colnames(validation) <- c("MSE", "RMSE", "r2")
rownames(validation) <- NULL
validation

ggplot(t, aes(x = rf.pred, y = rf.y)) +
  geom_bin2d(bins = 60) +
  geom_smooth(method = "lm") +
  geom_abline(intercept = 0, slope = 1, colour = "grey", linewidth = 1.2) +
  scale_fill_continuous(type = "viridis") +
  theme_bw() +
  scale_x_continuous(name = "Predicted value") +
  scale_y_continuous(name = "Observed value") +
  ggtitle("RF mean")

imp <- varImp(rf, scale = FALSE)
imp$Predictor <- rownames(imp)
rownames(imp) <- NULL
imp <- imp[order(imp[1], decreasing = TRUE), c(2, 1)]
colnames(imp)[2] <- "IncMSE"

impfig <- imp %>%
  mutate(Predictor = fct_reorder(Predictor, IncMSE)) %>%
  ggplot( aes(x=Predictor, y=IncMSE)) +
    geom_bar(stat="identity", fill="#f68060", alpha=.6, width=.4) +
    coord_flip() +
    xlab("") +
    ylab("% increase in MSE") +
    theme_bw()
    
impfig
```


## Spatial autocorrelation range

### SAR based on observations

The spatial dependence structure in the raw data is determined. Specifically, the distance (range) up to which observations are spatially autocorrelated is estimated with a variogram.

*Data need to be in an equidistant projection and (approximately) normally distributed.*

```{r spatial_autocorrelation_range_obs}
SAR_pr <- project(x = resp_centr, y = "+proj=eqdc +lon_0=12.5 +lat_1=59.5 +lat_2=77.5 +lat_0=68.5 +datum=WGS84 +units=m +no_defs")
SAR_pr <- as(SAR_pr, "Spatial")

if (lambda >  0){
  SAR_pr$SAR_cm_yr = SAR_pr$SAR_cm_yr ^ lambda
} else if (lambda == 0){
    SAR_pr$SAR_cm_yr = log(SAR_pr$SAR_cm_yr)
} else if (lambda <  0){
      SAR_pr$SAR_cm_yr = -1 * SAR_pr$SAR_cm_yr ^ lambda}

# automap package
vf <- autofitVariogram(SAR_cm_yr ~ 1, 
                       SAR_pr, 
                       #cutoff = 5000000
                       )
plot(vf)
sar <- vf$var_model$range[2]
```


## Creating spatial blocks

Spatial blocks and folds are created. The folds will be used in a spatial k-fold cross validation. The size of the blocks is determined by the spatial autocorrelation range.

Roberts et. al. (2017) suggest that blocks should be substantially bigger than the range of spatial autocorrelation (in model residual) to obtain realistic error estimates, while a buffer with the size of the spatial autocorrelation range would result in a good estimation of error.

*Should we increase the block size? This could be gauged by looking at the geographic distances plot below. The block size might be right, when sample-to-prediction and CV distances look similar.*

```{r spatial_blocks, warning=FALSE}
k <- 10 # Number of folds
m <- 0.35 # Multiplier applied to block size

spBlocks <- cv_spatial(x = SAR_pr,
                       k = k,
                       #hexagon = FALSE,
                       size = sar * m,
                       seed = 42,
                       progress = FALSE)
```


## Reshaping index

The output from the blocking step needs to be reshaped.

```{r reshape_index}
index_train <- list()
index_val <- list()
for (n in 1:spBlocks$k) {
  ft <- spBlocks[["folds_list"]][[n]][[-2]]
  fv <- spBlocks[["folds_list"]][[n]][[2]]
  index_train[[length(index_train)+1]] <- ft
  index_val[[length(index_val)+1]] <- fv
}
```


## Distances in geographic space including CV distances

```{r geogr_space_dist2, message=FALSE}
dist_geogr2 <- plot_geodist(st_as_sf(resp_centr), predictors,
                     cvfolds= index_val,
                     type = "geo",
                     unit="km",
                     showPlot = FALSE)

dist_geogr2$plot
dist_geogr2$plot + scale_x_log10()
```


## Model tuning

A Random Forest model is tuned. Predictor variables are finally selected in a forward feature selection approach and various values of the mtry parameter are tested in a spatial k-fold cross validation.

This step is time-consuming and memory-heavy. Therefore, only a subset of possible mtry values is tested. These are multiples of the default mtry values or the default values. 

The maximum number of iterations can be calculated upfront, based on the number of pre-selected predictors:

```{r max_iter}
factorial(length(sel_preds))/(factorial(2)*factorial(length(sel_preds)-2)) + sum(c((length(sel_preds)-2):1))
```


### Forward feature selection

The best combination of predictor variables (features) is found in a forward feature selection process.


```{r model_tuning}
nCores <- detectCores()
cl <- makePSOCKcluster(nCores - 1)
registerDoParallel(cl)

set.seed(42)

model <- ffs(seldata[sel_preds],
               seldata$SAR,
               metric = "Rsquared",
               method="qrf",
               what = 0.5,
               replace = FALSE,
               importance = TRUE,
               #tuneGrid = expand.grid(mtry =  round(floor(sqrt(length(sel_preds)))*c(1,1.5,2,3))),
               trControl = trainControl(method="CV",
                                        number = k,
                                        savePredictions = "final",
                                        index = index_train, 
                                        allowParallel = TRUE),
               verbose = TRUE)

stopCluster(cl)

model

sel_preds <- model$selectedvars
```


### FFS plot

Plot of R2 over the model runs.

```{r ffs_plot}
plot_ffs(model)
```


## Validation statistics

The validation results of the optimal RF model.

Note that these are the statistics based on the predicted values of the selected model. These differ slightly from the values from the tuning (above), which are the means of the k predictions based on the folds.

```{r validation_stats}
t <- data.frame(model$pred$pred, model$pred$obs)

validation <- data.frame(mse=numeric(), rmse=numeric(), r2=numeric())
validation[1,1] <- round(mse(t$model.pred.obs, t$model.pred.pred), 3)
validation[1,2] <- round(rmse(t$model.pred.obs, t$model.pred.pred), 3)
validation[1,3] <- round(cor(t$model.pred.obs, t$model.pred.pred)^2, 3)

colnames(validation) <- c("MSE", "RMSE", "r2")
rownames(validation) <- NULL
validation
```


## Validation plot

```{r validation_plot, message=FALSE}
ggplot(t, aes(x = model.pred.pred, y = model.pred.obs)) +
  geom_hex(bins = 60) +
  geom_smooth(method = "lm") +
  geom_abline(intercept = 0, slope = 1, colour = "grey", linewidth = 1.2) +
  scale_fill_continuous(type = "viridis") +
  theme_bw() +
  scale_x_continuous(name = "Predicted value") +
  scale_y_continuous(name = "Observed value")
       
```


## Variable importance

```{r variable_importance_plot}
imp <- varImp(model$finalModel)
imp$Predictor <- rownames(imp)
rownames(imp) <- NULL
imp <- imp[order(imp[1], decreasing = TRUE), c(2, 1)]
colnames(imp)[2] <- "IncMSE"
imp

impfig <- imp %>%
  mutate(Predictor = fct_reorder(Predictor, IncMSE)) %>%
  ggplot( aes(x=Predictor, y=IncMSE)) +
    geom_bar(stat="identity", fill="#f68060", alpha=.6, width=.4) +
    coord_flip() +
    xlab("") +
    ylab("% increase in MSE") +
    theme_bw()
    
impfig
```


## Distances in environmental space including CV distances

```{r env_space_dist2}
#dist_env2 <- plot_geodist(st_as_sf(resp_centr), predictors,
#                     type = "feature",
#                     cvfolds= index,
#                     variables = sel_preds,
#                     showPlot = FALSE)

#dist_env2$plot
#dist_env2$plot + scale_x_log10()
```



## Partial dependence

Partial dependence plots give a graphical depiction of the marginal effect of a variable on the response.

```{r partial_plots}
m2 <- model$finalModel
class(m2) <- "randomForest"

for (i in 1:length(sel_preds)) {
  partialPlot(x = m2, pred.data = seldata, x.var = sel_preds[i], main = "", xlab = sel_preds[i], ylab = paste0(resp_type, " (", resp_unit, ")"))
}
```


# Predict QRF model

## Predict SAR

```{r predict_sar}
preds <- stack(predictors[[sel_preds]])
SAR_med <- predict(preds, model$finalModel, what = 0.5)
SAR_p95 <- predict(preds, model$finalModel, what = 0.95)
SAR_p5 <- predict(preds, model$finalModel, what = 0.05)
SAR_pi90 <- SAR_p95 - SAR_p5
SAR_pir <- SAR_pi90 / SAR_med

hist(SAR_med, main = "", xlab = paste0(resp_type, " (", resp_unit, ")"))
```


## Area of Applicability

```{r aoa}
SAR_trainDI <- trainDI(model = model,
                        variables = sel_preds)
print(SAR_trainDI)

SAR_aoa <- aoa(newdata = predictors, 
                model = model,
                trainDI = SAR_trainDI,
                variables = sel_preds,
                )

plot(SAR_aoa)
```


## Plot results

```{r plot_results}
plot(SAR_med, main = paste0(resp_type, " (", resp_unit, ")"))
plot(SAR_pi90, main = "90% prediction interval")
plot(SAR_pir, main = "Prediction interval ratio")
plot(SAR_aoa$DI, main = "Dissimilarity index")
plot(SAR_aoa$AOA, main = "Area of applicability")

fr <- freq(SAR_aoa$AOA)
print(paste0("AOA = ", round(100*fr$count[2]/ sum(fr$count),2), "% of pixels"))
```


## Convert AOA from raster to polygon

```{r aoa_poly}
aoa_poly <- as.polygons(SAR_aoa$AOA, dissolve = TRUE)
plot(aoa_poly)

write_sf(st_as_sf(aoa_poly), dsn = "output", layer = paste0(resp_type, "_AOA_", Sys.Date()), driver = "ESRI Shapefile")
```


## Export geoTifs

```{r geotifs}
writeRaster(SAR_med, paste0("output/", resp_type, "_median_", Sys.Date(), ".tif"), overwrite = TRUE)
writeRaster(SAR_pi90, paste0("output/", resp_type, "_pi90_", Sys.Date(), ".tif"), overwrite = TRUE)
writeRaster(SAR_pir, paste0("output/", resp_type, "_pir_", Sys.Date(), ".tif"), overwrite = TRUE)
writeRaster(SAR_aoa$AOA, paste0("output/", resp_type, "_aoa_", Sys.Date(), ".tif"), overwrite = TRUE)
```


## Output a log file

```{r log}
sink(file = paste0("output/ModelLog_", Sys.Date(), ".txt"))
print("Selected Predictors")
sel_preds
model
print("Final Model")
paste0("MSE = ", validation[1,1])
paste0("RMSE = ", validation[1,2])
paste0("R2 = ", validation[1,3])
paste0("AOA = ", round(100*fr$count[2]/ sum(fr$count),2), "% of pixels")
sink()
```


# Finishing off

# Save QRF model

```{r save_model}
saveRDS(model, "qrfmodel.rds")
```


## Save session info

```{r save_session_info}
sessionInfo <- sessionInfo()
save(sessionInfo, file = "sessionInfo.Rdata")
rm("sessionInfo")
```


## Save global environment

```{r save_global_env}
save.image(file = "globEnv.RData")
```
